---
title: "Bike share Case Study"
author: "Bruno Schwartz"
date: "2024-02-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Prepare

### Libraries

```{r}
#tinytex::install_tinytex()
```


```{r}
#install.packages("tidyr")
#install.packages("readr")
#install.packages("dplyr")
#install.packages("lubridate")
#install.packages("skimr")
#install.packages("LaTeX")
```

```{r}
library(tidyr)
library(readr)
#library(janitor)
library(dplyr)
library(lubridate)
library(skimr)
library(ggplot2)
```

### Read the dataset

```{r}
Q12019 <- read_csv("Dataset/Divvy_Trips_2019_Q1.csv")
Q22019 <- read_csv("Dataset/Divvy_Trips_2019_Q2.csv")
Q32019 <- read_csv("Dataset/Divvy_Trips_2019_Q3.csv")
Q42019 <- read_csv("Dataset/Divvy_Trips_2019_Q4.csv")
Q12020 <- read_csv("Dataset/Divvy_Trips_2020_Q1.csv")
```

```{r}
str(Q12019)
```

```{r}
glimpse(Q12019)
glimpse(Q42019)
```

```{r}
glimpse(Q22019)
glimpse(Q32019)
```


```{r}
head(Q22019)
```

### Rename of columns

```{r}
# Conversion to date time datatype
colnames(Q12019)
colnames(Q22019)
colnames(Q32019)
colnames(Q42019)
colnames(Q12020)
```

```{r}
Q22019 <- Q22019 %>% 
  rename(
    trip_id = "01 - Rental Details Rental ID",
    start_time = "01 - Rental Details Local Start Time",
    end_time = "01 - Rental Details Local End Time",
    bikeid = "01 - Rental Details Bike ID",
    tripduration = "01 - Rental Details Duration In Seconds Uncapped",
    from_station_id = "03 - Rental Start Station ID",
    from_station_name = "03 - Rental Start Station Name",
    to_station_id = "02 - Rental End Station ID",
    to_station_name = "02 - Rental End Station Name",
    usertype = "User Type",
    gender = "Member Gender",
    birthyear = "05 - Member Details Member Birthday Year"
    )
```

```{r}
colnames(Q22019)
```
```{r}
Q12020 <- Q12020 %>% 
  rename(
    trip_id = ride_id,
    start_time = started_at,
    end_time = ended_at,
    from_station_id = start_station_id,
    from_station_name = start_station_name,
    to_station_id = end_station_id,
    to_station_name = end_station_name,
    usertype = member_casual
    )
```

```{r}
colnames(Q12020)
```

### Conversion of data to datetime

```{r}
# Conversion to date time datatype
Q12019$start_time <- mdy_hm(Q12019$start_time)
Q12019$end_time <- mdy_hm(Q12019$end_time)
```

### Analyze the columns

```{r}
# Gives us a pretty comprehensive summary of a dataset.
#skim_without_charts(Q12019)
skim(Q12019)
```

### Combine the dataframes in one

```{r}
# Remove columns
Q12020 <- Q12020 %>%
  select(-rideable_type, -start_lat, -start_lng, -end_lat, -end_lng)
Q12019 <- Q12019 %>%
  select(-bikeid)
Q22019 <- Q22019 %>%
  select(-bikeid)
Q32019 <- Q32019 %>%
  select(-bikeid)
Q42019 <- Q42019 %>%
  select(-bikeid)
```


```{r}
# Add columns
Q12020$gender = NA
Q12020$birthyear = NA
#Calculate the value of the trip duration
Q12020 <- Q12020 %>% 
  mutate(tripduration = as.numeric(difftime(end_time, start_time, units = "secs")))
```

```{r}
# Merge the dataframes
Cyclistic <- rbind(Q12019, Q22019, Q32019, Q42019, Q12020)
```


## Process

```{r}
glimpse(Cyclistic)
```

### Assess the trip duration variable:

```{r}
#Check negative values
Cyclistic %>%
  filter(tripduration<0)
```

```{r}
#Check empty values
Cyclistic %>%
  filter(tripduration=NA)
```

Remove the negative values of the tripduration column
```{r}
Cyclistic <- Cyclistic %>%
  filter(tripduration>=0)
```


### Calculate the weekday: day_of_week

```{r}
head(weekdays(Cyclistic$start_time))
```

```{r}
Cyclistic <- Cyclistic %>% 
  mutate(day_of_week = weekdays(start_time))
```

```{r}
head(Cyclistic)
```


### Calculate the starting hour: start_hour

```{r}
head(hour(Cyclistic$start_time),50)
head(Cyclistic$start_time,50)
```

```{r}
Cyclistic <- Cyclistic %>% 
  mutate(start_hour = hour(start_time))
```

```{r}
Cyclistic %>%
  select (start_time, start_hour)
```

### Replace the values of column usertype

```{r}
unique(Cyclistic$usertype)
```

```{r}
Cyclistic$usertype[Cyclistic$usertype == "member"] <- "Subscriber"
Cyclistic$usertype[Cyclistic$usertype == "casual"] <- "Customer"
```

```{r}
unique(Cyclistic$usertype)
```

## Analyze

```{r}
skim(Cyclistic)
```

Unusual ride_length values (excessively high):
```{r}
Cyclistic %>%
  filter(tripduration/3600 > 1000)
```

### Number of rides per day and usertype 

```{r}
Cyclistic$day_of_week <- factor(Cyclistic$day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
ggplot(Cyclistic, aes(x=day_of_week, fill=usertype)) + 
  geom_bar(position="dodge") +
  labs(x="Day of the week", y="Number of rides") + 
  scale_y_continuous(label=scales::comma)
```

### Average trip duration (in hours) per usertype

```{r}
Cyclistic_filtered <- Cyclistic %>%
  select(start_time, usertype, tripduration, day_of_week)
# Convert the variable from Datetime to Date
Cyclistic_filtered$start_time <- as.Date(Cyclistic_filtered$start_time)
```

```{r}
Cyclistic_Subscriber <- Cyclistic_filtered %>%
  filter(usertype=="Subscriber") %>%
  group_by(start_time, usertype, day_of_week) %>%
  summarise(mean_tripduration = mean(tripduration)/3600)
```

```{r}
Cyclistic_Customer <- Cyclistic_filtered %>%
  filter(usertype=="Customer") %>%
  group_by(start_time, usertype, day_of_week) %>%
  summarise(mean_tripduration = mean(tripduration)/3600)
```

```{r}
ggplot(Cyclistic_Subscriber, aes(x=start_time, y=mean_tripduration)) + 
  geom_line() +
  labs(x="Date", y="Average trip duration (in hours)")
```

```{r}
ggplot(Cyclistic_Customer, aes(x=start_time, y=mean_tripduration)) + 
  geom_line() +
  labs(x="Date", y="Average trip duration (in hours)")
```

```{r}
Cyclistic_filtered <- Cyclistic %>%
  select(usertype, tripduration, day_of_week)
```

```{r}
Cyclistic_day_Subscriber <- Cyclistic_filtered %>%
  filter(usertype=="Subscriber") %>%
  group_by(day_of_week) %>%
  summarise(mean_tripduration = mean(tripduration)/3600)
```

```{r}
Cyclistic_day_Customer <- Cyclistic_filtered %>%
  filter(usertype=="Customer") %>%
  group_by(day_of_week) %>%
  summarise(mean_tripduration = mean(tripduration)/3600)
```

```{r}
Cyclistic_day_Subscriber
```

```{r}
Cyclistic_day_Customer
```

### Number of rides per hour and usertype 

```{r}
ggplot(Cyclistic, aes(x=start_hour, fill=usertype)) + 
  geom_bar(position="dodge") +
  labs(x="Starting hour", y="Number of rides") + 
  scale_y_continuous(label=scales::comma)
```



```{r}
ggplot(Cyclistic, aes(x=start_hour, fill=usertype)) + 
  geom_bar(position="dodge") +
  facet_wrap(~day_of_week) +
  labs(x="Starting hour", y="Number of rides") + 
  scale_y_continuous(label=scales::comma)
```
